Water Management

### 2020-02-27 Paper/paper.tex
add TODO
fix a bunch of TODOs

In this work, we want to examine the variations in use of water from a (hypothetical) water reservoir.
One of the most important uses of surface water is the irrigation of fields for food production. However, as farming and food production are private enterprises and lakes are public goods, interests of different parties have to be considered. 
The water reservoir to be considered has two stakeholders: farmers, who want to use the water for irrigation purposes and citizens, who want to use the lake for recreational activities. 
Though looking at just two agents is a vast simplification compared to real allocation problems, one can easily add more stakeholders to the model, as long as there is enough computational power.
The sequence of events for every period is displayed in \ref{fig:Sequence}. %TODO Die Figure ist im PDF zu weit weg gerutscht! 
At the beginning of every year (Step 1) we have to decide how much of the water from the reservoir is used for the irrigation of fields, which benefits the farmers' harvest.\\\\
The farmers utility function follows the form

@@

At the end of the year, water in the reservoir is replenished by rainfall (Step 3), which we can assume to be randomly distributed following a lognormal distribution \citep{oosterbaan1994frequency}: %TODO cite steht über den Rand hinaus
\begin{equation}
	r \sim lognormal(\mu, \sigma^2)
\end{equation}

@@

In this work, we will determine an optimal irrigation policy that maximizes the combined utility of farmers and citizens over $T$ years by using dynamic programming in combination with a stochastic simulation of the rainfall. 
In the main part we will give a brief summary of the underlying theory and the optimization problem. 
After that we will present the results and our conclusions for the original model and its modifications, as well as possible shortfalls of our analysis. %TODO statt "possible shortfalls" besser "possible enhancements" schreiben? (klingt positiver?)
Finally, we will give an outlook for further possible extensions.

\subsubsection{Modification 1: Variation of the Discount Factor}
In this modification, we're interested in studying the effects of different betas on the optimal irrigation policy. 
The discount factor takes into account that agents in a simple two-stage model often value current consumption over future consumption.
It expresses how much one unit of future utility is worth in terms of current utility and therefore makes them mathematically comparable \citep{kruschwitz2014investitionsrechnung}. %TODO wirklich deutsche Citation verwenden? 
We take a look at the discount factors from X to Y. %TODO tatsächliche Grenzen einfügen

\subsubsection{Modification 2: Increasing the Volatility of Rainfalls}
In order to take into account expected future impacts of climate change \citep{guhathakurta2011impact}, we consider the effects of increased volatility of precipitation. %TODO cite steht über den Rand hinaus
We do so by setting $\sigma$ to 2. %TODO Das ist noch nicht ganz richtig

@@

\subsubsection{Modification 4: Taking Evaporation into Account} \label{sec:intro-mod4}
The evaporation of water from a reservoir is the most important natural "outflow" to be considered when it comes to managing water levels \citep{tanny2008evaporation}. 
To determine a more realistic optimal irrigation strategy, we should therefore consider evaporation effects in our model. 
The dataset that we use for the real rainfall also contains many other weather parameters.
The dataset from \citep{kaggle:2019} does not contain evaporation information directly. However, with the mean temperature $T$ in °C, the elevation $h$ in m, the latitude $A$ in °S and the dew point $T_d$ in °C we can calculate an estimate for the evaporation of water in mm per day for a specific location, using the formula \citep{linacre1977simple}

@@

%TODO sind diese beiden Sätze so schön? oder nötig?
Note that all modifications, except Modification 4, are based on the original model without any modifications. 
Modification 4 is an extension of Modification 3.

@@

A common technique to solve optimization or combinatorial problems is called Dynamic Programming. 
It is based on the idea of dividing problems into smaller subproblems, solving these recursively and caching intermediate results (\textit{memoization}). % Nicht eher "memoRization"?
In his paper "The Theory Of Dynamic Programming" from 1957, Richard Bellman aimed to provide a solution to multi-stage-decision processes. 
He defines a \textit{policy} as "a sequence of decisions" (\cite{bellman:1957}, p.2), and an \textit{optimal policy} as "a policy, which is most advantageous according to some preassigned criterion" (\cite{bellman:1957}, p.2). 
This is reflected in Bellman's Principle of Optimality: "An optimal policy has the property that whatever initial state and initial decisions are, the remaining decisions must constitute an optimal policy with regard to the state resulting from the first decisions" (\cite{bellman:1957}, p.2).

@@

So the value of the control variable to be set is the one that maximizes both the utility of the current state plus (recursively) the value $V$ of the next state $s'$ which is discounted by a factor of $\beta$. 
%TODO hier noch auf die Definition von beta eingehen? Was ist beta und warum ist das relevant?
\subsection{Implementation}
\subsubsection{Computation of the value function}

@@

which equals to finding the column $aux$ that maximizes the expected value over all states.
We repeat this step over and over again until the our value function $V$ converges.
In our case the function converges, if the norm of the difference between $V'$ and $V$ falls below a certain tolerance value \\
% TODO Fehlen hier nicht noch ein paar Sätze dazu wie wir die optimal irrigation policy extrahieren?

@@

%------------------------------------------------
\newpage
\section{Results}
This chapter will discuss the outcomes and results of the implemented functionality in MATLAB.
The discussion will be started on the basic questions asked by the project description.
Furthermore we made several modifications to the task which will be discussed additionally.
All results will be visualized by plots made within our simulation in MATLAB.
\subsection{Optimal irrigation policy}
When examining what an optimal irrigation policy could be, the first question to be asked is how the aggregated value function of both recreational users and farmers looks like. 
The value function as posed in \ref{eq:value-function} contains the maximum possible (expected) utility that can be gained by irrigating a certain amount of water under a given water level in the reservoir. 
Figure \origref{fig:value-function} shows the value function calculated for all possible water levels between the allowed boundaries of $0$ and $7$. 
\begin{figure}[ht]
	\includegraphics[width=1\textwidth]{figures/value_function.eps}
	\caption{Value Function}
	\label{fig:value-function}
\end{figure}
\newline
Every value along the possible water levels is mapped to an irrigation amount that optimizes the expected value for all parties in the given situation. 
The resulting optimal irrigation policy is shown in \ref{fig:optimal-irrigation-policy}. 
As the plot shows, a higher water level leads to a higher proposed irrigation amount. There are two particularities that can be found within the graph:
\begin{itemize}
	\item Due to the effect that an almost empty reservoir has a very low use to recreational users, the optimal irrigation strategy suggests a very low irrigation amount whose gradient increases with increasing water level.
	\item As long as the discount factor $\beta$ is greater than $0$, the water within the reservoir has a value in the next period for both recreational users and farmers. Due to this effect, saving waters makes sense and leads to a decreasing gradient for the proposed irrigation amount with increasing water level.
\end{itemize}
\begin{figure}[ht]
	\includegraphics[width=1\textwidth]{figures/OptimalIrrigationPolicy.eps}
	\caption{Optimal Irrigation Policy}
	\label{fig:optimal-irrigation-policy}
\end{figure}
\newpage
\subsection{Steady State}
In systems theory, a system or a process is in a steady state, if the variables which define the behavior of the system are unchanging in time \citep{gagniuc2017markov}. 
The variable defining the behavior of our system is the water level. 
Thus, we have to calculate the change in water level for each period. 
Due to the fact that the water level highly depends on the randomly distributed rainfalls (that fill the reservoir at end of each period), it is not possible to find a steady state in which the water level does not change \textit{at all}. 
In this context, we define a steady state to be a state, where the mean water level of the last 200 periods does not change more than $1*10^{-7}$ from one period to the next. %TODO ist dieser Wert noch richtig? Im Code steht im Moment steadyStateTolerance = 0.0001;
A exemplary graph of the mean in every period and the difference to its predecessor is shown in \ref{fig:steadyState}.
Once a period is found where the mean does not change more than the given tolerance, the mean of that period is defined as the steady state of the water level. %TODO hier "moving average" schreiben?
The water level found to be a steady state and the period in which this condition occurs changes from one simulation to the next, since the rainfall is generated randomly following the given distribution.

@@

\subsection{Steady State Probability}
As just mentioned, the steady state changes from simulation to simulation. 
In order to find out how the steady states are distributed and what the probability of their occurrences are, we implement a Monte Carlo simulation. 
To do so, we evaluate the forward simulation 1000 times and put the resulting water levels into bins with the size of 0.1 each. 
We then counted the occurrences, normalized them and calculated the corresponding probability.
The result is put in a histogram and shown in \ref{fig:steadyStateMonteCarlo}.
 %TODO ich (martin) würde vorschlagen, einfach zu schreiben ... "1000 times and calculate a relative frequency distribution" (ist hier erklärt: https://www.sciencedirect.com/topics/mathematics/relative-frequency-distribution)
The highest possibility for a steady state is around 4.2. 
The curve resembles a normal distribution with a standard deviation of $\sigma=0.37$ and peaks at a water level around $4.2$.

@@

\subsection{Effect of the Discount Factor on the Optimal Irrigation Policy}
When altering the discount factor $\beta$ in a range from 0 to 1, the optimal irrigation policy changes. %TODO auch hier die tatsächlichen Werte anpassen?!
For small water levels of the reservoir, the optimal irrigation amount is rather insensitive to the discount factor. 
A small $\beta$ ensures that high water levels at the end of one period are only available to a small extent in the next period. 
%TODO Finde die Formulierung nicht so gut: Sie sind ja weiterhin voll verfügbar, aber eben in Zukunft weniger wert. Bzw das könnte man diskutieren. 
This leads to an increasing gap between irrigation amounts over $\beta$ with increasing water level since conserving water does not increase the usage as much. 
One can say that a high $\beta$ prefers a long-term policy while a smaller $\beta$ leads to a rather short-term policy.
%TODO ich (martin) würde diesen absatz so schreiben "A smaller $\beta$ allows more water to be used for irrigation today, optimizing the short-term utility. On the other hand, a higher $\beta$ increases the today's valuation of future utility, and therefore reduces the amount of water to be consumed (irrigated) today. "
\begin{figure}[ht]
	\includegraphics[width=1\textwidth]{figures/optimum-policy-over-beta.eps}
	\caption{Optimal irrigation policy as a function of the discount factor $\beta$}
	\label{fig:optimal-irrigation-beta}
\end{figure}
\newpage
\subsection{Increased Volatility of the Rainfalls} %TODO überall use durch utility ersetzen!
More volatile rainfalls means that rainfalls are having a higher standard deviation.
Since we want to compare two rainfall distributions of the same mean, the median of the distribution has to be changed. 
Originally, we assumed $\mu=0$ and $\sigma=1$. Now, we choose $\sigma=2$ and thus have to set $\mu=-1.5$, in order to retain the same mean (see \ref{eq:rainfall}). %TODO ich (martin) hab diesen Absatz so umgeschrieben, einverstanden? Vorher stand hier insbesondere auch (fälschlicherweise?!) $\sigma^2=2$: 
%Since we want to compare two rainfall distributions of the same mean, the median of the distribution has to be changed.
%Using a $\sigma^2=2$ includes setting $\mu=-1.5$ in order for the mean to stay the same.
%(See \ref{eq:rainfall} for reference)

@@

\ref{fig:evaporation-factors} shows the total utility and the water level over time, using scaling factors of $0, 1, 2, 2.3$.\footnote{The expected value of the evaporation should not exceed the expected value of the rainfall, as this would make it impossible to calculate the value function. The expected value of the evaporation with a scaling factor of $2.3$ is just below the expected value of the rainfall. } %TODO Fußnote gut so?
We can see that higher amounts of evaporation not only lead to water levels, but also to significantly lower utilities for the recreational users and the farmers. 
Especially when the reservoir completely dries out, the total utility approaches $-\infty$. This happens in periods $7,62,63,64$ and $65$ when applying a factor of $2.3$ to the evaporation.

@@

%TODO content

@@

Running our forward simulation 500 times, we could see that it seems to follow a normal distribution with a mean of around 4.2 and a standard deviation of 0.37. % TODO: StdDev auch im Ergebnisteil erwähenen? 

@@

>>>>

In this work, we want to examine the variations in use of water from a (hypothetical) water reservoir.
One of the most important uses of surface water is the irrigation of fields for food production. However, as farming and food production are private enterprises and lakes are public goods, interests of different parties have to be considered. 
The water reservoir to be considered has two stakeholders: farmers, who want to use the water for irrigation purposes and citizens, who want to use the lake for recreational activities. 
Though looking at just two agents is a vast simplification compared to real allocation problems, one can easily add more stakeholders to the model, as long as there is enough computational power.
The sequence of events for every period is displayed in \ref{fig:Sequence}.
At the beginning of every year (Step 1) we have to decide how much of the water from the reservoir is used for the irrigation of fields, which benefits the farmers' harvest.\\\\
The farmers utility function follows the form

@@

At the end of the year, water in the reservoir is replenished by rainfall (Step 3), which we can assume to be randomly distributed following a lognormal distribution \citep{oosterbaan1994frequency}: 
\begin{equation}
	r \sim lognormal(\mu, \sigma^2)
\end{equation}

@@

In this work, we will determine an optimal irrigation policy that maximizes the combined utility of farmers and citizens over $T$ years by using dynamic programming in combination with a stochastic simulation of the rainfall. 
In the main part we will give a brief summary of the underlying theory and the optimization problem. 
After that we will present the results and our conclusions for the original model and its modifications, as well as possible enhancements of our analysis.
Finally, we will give an outlook for further possible extensions.

\subsubsection{Modification 1: Variation of the Discount Factor}
In this modification, we're interested in studying the effects of different betas on the optimal irrigation policy. 
The discount factor takes into account that agents in a simple two-stage model often value current consumption over future consumption.
It expresses how much one unit of future utility is worth in terms of current utility and therefore makes them mathematically comparable \citep{kruschwitz2014investitionsrechnung}. 
We take a look at the discount factors from 0.05 to 0.95.

\subsubsection{Modification 2: Increasing the Volatility of Rainfalls}
In order to take into account expected future impacts of climate change \citep{guhathakurta2011impact}, we consider the effects of increased volatility
We do so by setting $\sigma=2$ and $\mu=-1.5$.

@@

\subsubsection{Modification 4: Taking Evaporation into Account} \label{sec:intro-mod4}
The evaporation of water from a reservoir is the most important natural "drain" to be considered when it comes to managing water levels \citep{tanny2008evaporation}. 
To determine a more realistic optimal irrigation strategy, we should therefore consider evaporation effects in our model. 
The dataset that we use for the real rainfall also contains many other weather parameters.
The dataset from \citep{kaggle:2019} does not contain evaporation information directly. However, with the mean temperature $T$ in °C, the elevation $h$ in m, the latitude $A$ in °S and the dew point $T_d$ in °C we can calculate an estimate for the evaporation of water in mm per day for a specific location, using the formula \citep{linacre1977simple}

@@

@@

A common technique to solve optimization or combinatorial problems is called Dynamic Programming. 
It is based on the idea of dividing problems into smaller subproblems, solving these recursively and caching intermediate results (\textit{memoization}). % Nicht eher "memoRization"?
In his paper "The Theory Of Dynamic Programming" from 1957, Richard Bellman aimed to provide a solution to multi-stage-decision processes. 
% TODO unbedingt noch Bellman als Quelle hinzufügen!!!
He defines a \textit{policy} as "a sequence of decisions" (\cite{bellman:1957}, p.2), and an \textit{optimal policy} as "a policy, which is most advantageous according to some preassigned criterion" (\cite{bellman:1957}, p.2). 
This is reflected in Bellman's Principle of Optimality: "An optimal policy has the property that whatever initial state and initial decisions are, the remaining decisions must constitute an optimal policy with regard to the state resulting from the first decisions" (\cite{bellman:1957}, p.2).

@@

So the value of the control variable to be set is the one that maximizes both the utility of the current state plus (recursively) the value $V$ of the next state $s'$ which is discounted by a factor of $\beta$. 
%TODO hier noch auf die Definition von beta eingehen? Was ist beta und warum ist das relevant?
\subsection{Implementation}
%TODO theoretisch ist es sehr unschön ein Kapitel ohne Text aufzumachen. Ist aber nur Makulatur
\subsubsection{Computation of the value function}

@@

which equals to finding the column $aux$ that maximizes the expected value over all states.
We repeat this step over and over again until the our value function $V$ converges.
In our case the we define the function to be converging, if the norm of the difference between $V'$ and $V$ falls below a certain tolerance value.
Summarizing the optimal irrigation amounts for all possible water levels into one function results in the optimal irrigation policy.

@@

%------------------------------------------------
\newpage
\section{Results}
This chapter will discuss the outcomes and results of the implemented functionality in MATLAB.
The discussion will be started on the basic questions asked by the project description.
Furthermore we made several modifications to the task which will be discussed additionally.
All results will be visualized by plots made within our simulation in MATLAB.
\subsection{Optimal irrigation policy}
When examining what an optimal irrigation policy could be, the first question to be asked is how the aggregated value function of both recreational users and farmers looks like. 
The value function as posed in \ref{eq:value-function} contains the maximum possible (expected) utility that can be gained by irrigating a certain amount of water under a given water level in the reservoir. 
Figure \origref{fig:value-function} shows the value function calculated for all possible water levels between the allowed boundaries of $0$ and $7$. 
\begin{figure}[ht]
	\includegraphics[width=1\textwidth]{figures/value_function.eps}
	\caption{Value Function}
	\label{fig:value-function}
\end{figure}
\newline
Every value along the possible water levels is mapped to an irrigation amount that optimizes the expected value for all parties in the given situation. 
The resulting optimal irrigation policy is shown in \ref{fig:optimal-irrigation-policy}. 
As the plot shows, a higher water level leads to a higher proposed irrigation amount. There are two particularities that can be found within the graph:
\begin{itemize}
	\item Due to the effect that an almost empty reservoir has a very low use to recreational users, the optimal irrigation strategy suggests a very low irrigation amount whose gradient increases with increasing water level.
	\item As long as the discount factor $\beta$ is greater than $0$, the water within the reservoir has a value in the next period for both recreational users and farmers. Due to this effect, saving waters makes sense and leads to a decreasing gradient for the proposed irrigation amount with increasing water level.
\end{itemize}
\begin{figure}[ht]
	\includegraphics[width=1\textwidth]{figures/OptimalIrrigationPolicy.eps}
	\caption{Optimal Irrigation Policy}
	\label{fig:optimal-irrigation-policy}
\end{figure}
\newpage
\subsection{Steady State}
In systems theory, a system or a process is in a steady state, if the variables which define the behavior of the system are unchanging in time \citep{gagniuc2017markov}. 
The variable defining the behavior of our system is the water level. 
Thus, we have to calculate the change in water level for each period. 
Due to the fact that the water level highly depends on the randomly distributed rainfalls (that fill the reservoir at end of each period), it is not possible to find a steady state in which the water level does not change \textit{at all}. 
In this context, we define a steady state to be a state, where the mean water level of the last 200 periods does not change more than $1*10^{-4}$ from one period to the next. 
A exemplary graph of the mean in every period and the difference to its predecessor is shown in \ref{fig:steadyState}.
Once a period is found where the mean does not change more than the given tolerance, the mean of that period is defined as the steady state of the water level. 
The water level found to be a steady state and the period in which this condition occurs changes from one simulation to the next, since the rainfall is generated randomly following the given distribution.

@@

\subsection{Steady State Probability}
As just mentioned, the steady state changes from simulation to simulation. 
In order to find out how the steady states are distributed and what the probability of their occurrences are, we implement a Monte Carlo simulation. 
To do so, we evaluate the forward simulation 1000 times and put the resulting water levels into bins with the size of 0.1 each. 
We then counted the occurrences, normalized them and calculated the corresponding probability.
The result is put in a histogram and shown in \ref{fig:steadyStateMonteCarlo}.
The highest possibility for a steady state is around 4.2. 
The curve resembles a normal distribution with a standard deviation of $\sigma=0.37$ and peaks at a water level around $4.2$.

@@

\subsection{Effect of the Discount Factor on the Optimal Irrigation Policy}
When altering the discount factor $\beta$ in a range from 0.05 to 0.95, the optimal irrigation policy changes. 
For small water levels of the reservoir, the optimal irrigation amount is rather insensitive to the discount factor. 
A small $\beta$ ensures that high water levels at the end of one period only keep their value only to a small extent in the next period.  
This leads to an increasing gap between irrigation amounts over $\beta$ with increasing water level since conserving water does not increase the usage as much. 
A smaller $\beta$ allows more water to be used for irrigation today, optimizing the short-term utility. On the other hand, a higher $\beta$ increases the today's valuation of future utility, and therefore reduces the amount of water to be consumed (irrigated) today.
\begin{figure}[ht]
	\includegraphics[width=1\textwidth]{figures/optimum-policy-over-beta.eps}
	\caption{Optimal irrigation policy as a function of the discount factor $\beta$}
	\label{fig:optimal-irrigation-beta}
\end{figure}
\newpage
\subsection{Increased Volatility of the Rainfalls} 
More volatile rainfalls means that rainfalls are having a higher standard deviation.
Since we want to compare two rainfall distributions of the same mean, the median of the distribution has to be changed. 
Originally, we assumed $\mu=0$ and $\sigma=1$. Now, we choose $\sigma=2$ and thus have to set $\mu=-1.5$, in order to retain the same mean (see \ref{eq:rainfall}).

@@

\ref{fig:evaporation-factors} shows the total utility and the water level over time, using scaling factors of $0, 1, 2, 2.3$.\footnote{The expected value of the evaporation should not exceed the expected value of the rainfall, as this would make it impossible to calculate the value function. The expected value of the evaporation with a scaling factor of $2.3$ is just below the expected value of the rainfall. }
We can see that higher amounts of evaporation not only lead to water levels, but also to significantly lower utilities for the recreational users and the farmers. 
Especially when the reservoir completely dries out, the total utility approaches $-\infty$. This happens in periods $7,62,63,64$ and $65$ when applying a factor of $2.3$ to the evaporation.

@@

@@

Running our forward simulation 500 times, we could see that it seems to follow a normal distribution with a mean of around 4.2 and a standard deviation of 0.37.

@@

% TODO "whatever that may be" klingt echt merkwürdig
